---
title: "Charlottesville Regional Equity Dashboard Update Workflow"
output: html_document
---
Set up for RMD file
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
Additional packages may be included in other R script sources.
```{r}
# Libraries
library(tidycensus)
library(tidyverse)
library(sf)
library(janitor)
library(readxl)
```

## Set variables
```{r}
# Current year for dashboard release
dyr <- 2026
# Most current ACS year
acsyr <- 2024

# County FIPS codes
# BRHD: Albemarle, Charlottesville, Fluvanna, Greene, Louisa, Nelson
fips_codes <- c("003", "540", "065", "079", "109", "125")

# Update to most recent TIGER/Line Shapefiles
# https://www.census.gov/geographies/mapping-files.html
tigeryr <- 2025

```

## Set file paths for saving data
These file paths assume the current working directory is the directory where this parent file lives. Check with `getwd()` and set with `setwd()`.
```{r}
# Check and create data_sources folder if does not exist
if (!dir.exists("data")){
  dir.create("data", recursive = TRUE)}

# File path for data folder
filepath_data <- "data/2026_update/"

# File path for final app data
filepath_app <- "cville-region/www/"

```

# Data Sources

## ACS 5-yr survey data

### Set Census API Key
A key can be obtained from <http://api.census.gov/data/key_signup.html>. This typically only needs to be set once: `install = TRUE`, will install the key in your .Renviron file for use in future sessions. Do not save this file in a public or shared location with your individual API key included in the code below.
```{r}
# If API key has not been previously installed, add API key inside the quotes and uncomment the line below to run. 
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

### Review ACS table codes and variables
Review ACS table variables if needed.
```{r}
# Uncomment below as needed. 
# vars_acs <- load_variables(acsyr, "acs5", cache = TRUE)
# vars_subject <- load_variables(acsyr, "acs5/subject", cache = TRUE)
# vars_profile <- load_variables(acsyr, "acs5/profile", cache = TRUE)
```

### Get and Clean ACS5 data

* Check variable codes for DP05 table race/ethnicity in `vars_profile` above (often changes).
```{r, results='hide'}
source('scripts/acs5-county-data.R')
source('scripts/acs5-tract-data.R')
source('scripts/acs5-blockgroup-data.R')
```

### Check
```{r, results='hide'}
county_data %>% select_at(vars(ends_with("E"))) %>% summary()
tract_data %>% select_at(vars(ends_with("E"))) %>% summary()
blkgrp_data %>% select_at(vars(ends_with("E"))) %>% summary()
```

## Additional county and tract-level data

### County-level ALICE
* Get Virginia ALICE data: https://www.unitedforalice.org/county-reports/virginia

Before running script, update url and filename for current ALICE data
```{r}
source('scripts/alice-county-data.R')
```

### County-level life expectancy
* Life Expectancy Estimates: https://www.countyhealthrankings.org/explore-health-rankings/virginia/data-and-resources

Before running script, update url and filename for current County Health Rankings life expectancy data
```{r}
source('scripts/life-exp-county-data.R')
```

### County and Tract-level Segregation data
* Segregation measures (from ACS data, but with more derivation)

```{r}
source('scripts/seg-county-tract-data.R')
```

## Magisterial District geometry data
```{r}
source('scripts/magdist-data.R')

# Check
ggplot() + 
  geom_sf(data=mcd_sf, color="black", fill="white", size=0.25)
```

## School District data
From NCES: https://nces.ed.gov/programs/edge/Geographic/SchoolLocations

Update urls before running. 
For Private/Public School URLS:
(1) Go to: https://nces.ed.gov/programs/edge/Geographic/SchoolLocations
(2) Find most recent year for public and private schools
(3) Right click each file link and select `copy link`
(4) Paste results into associated url below

For School Boundaries:
(1) Go to: https://nces.ed.gov/programs/edge/SABS
(2) Find most recent year for school boundaries
(3) Right click Single Shapefile link and select `copy link`
(4) Paste results into associated url below
```{r}
# public schools - updated public schools as of 4/2026 are from 2024-2025 school year
urlpublic <- "https://nces.ed.gov/programs/edge/data/EDGE_GEOCODE_PUBLICSCH_2425.zip"

# private schools - updated private schools as of 4/46 are from 2023/24 school year
urlprivate <- "https://nces.ed.gov/programs/edge/data/EDGE_GEOCODE_PRIVATESCH_2324.zip"

# School boundaries - As of 4/26, the school boundaries are still from 2015-2016
# These take a long time to download - comment out in script if you did this in a previous year and still have the data
urlboundaries <- "https://nces.ed.gov/programs/edge/data/SABS_1516.zip"

source('scripts/schools-data.R')
```

## Parks Data - OpenStreet Map

```{r}
source('scripts/parks-osm-data.R')

# view
leaflet(cville_bounds) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(weight = 1, fill = 0, color = "black") %>% 
  addCircles(data = st_collection_extract(parks, "POINT"), color = "blue",
             popup = st_collection_extract(parks, "POINT")$name) %>% 
  addPolygons(data = st_collection_extract(parks, "POLYGON"), color = "orange",
              popup = st_collection_extract(parks, "POLYGON")$name)

```


# Prepare data for Dashboard

## Update Metadata

### Data Sources
Review the below Excel workbook `prettytable_data_sources.xlsx` saved in `scripts/` and update the following:
* Source column in each sheet to reflect current ACS 5-year survey and Accessed Data (use Find & Replace)
* In County sheet, update source and about info for Life Expectancy data

```{r}
# Read in data sources workbook sheets
metadata <- read_excel("scripts/prettytable_data_sources.xlsx", sheet = "acs_county")
metadata$label <- metadata$description

# metadata$label <- str_to_title(metadata$description)

# metadata_tract <- read_excel("scripts/prettytable_data_sources.xlsx", sheet = "acs_tract")
# metadata_tract$label <- str_to_title(metadata_tract$description)
# 
# metadata_blkgrp <- read_excel("scripts/prettytable_data_sources.xlsx", sheet = "acs_blockgroup")
# metadata_blkgrp$label <- str_to_title(metadata_blkgrp$description)
```

### Regional Tractnames
```{r, results='hide'}
# Read in Regional Tract names csv
tractnames <- read_csv("scripts/regional_tractnames.csv")

# Clean up tractnames
tractnames <- tractnames %>%
  mutate(GEOID = as.character(GEOID)) %>% 
  rename(coname = locality,
         locality = locality_num)
```

## Load data 
Uncomment if not in current environment
```{r}
# ACS Data
county_data <- read_rds(paste0(filepath_data, "acs5_county.RDS"))
tract_data <- read_rds(paste0(filepath_data, "acs5_tract.RDS"))
blkgrp_data <- read_rds(paste0(filepath_data, "acs5_blockgroup.RDS"))

# Additional data
alice <- read_rds(paste0(filepath_data, "alice.RDS"))
life_exp <- read_rds(paste0(filepath_data, "county_life_exp.RDS"))
seg_county <- read_rds(paste0(filepath_data, "seg_county.RDS"))
seg_tract <- read_rds(paste0(filepath_data, "seg_tract.RDS"))

# Points and Polygons
mcd_sf <- st_read(paste0(filepath_data, "mcd_sf.geojson"))
schools_sf <- st_read(paste0(filepath_data, "schools_sf.geojson"))
sabselem_sf <- st_read(paste0(filepath_data, "sabselem_sf.geojson"))
sabshigh_sf <- st_read(paste0(filepath_data, "sabshigh_sf.geojson"))
parks_sf <- st_read(paste0(filepath_data, "parks_OSM_sf.geojson"))
```

## Merge data
```{r}
# Merge tract & block group data ----

# add segregation measures by tract
tract_data <- tract_data %>% 
  left_join(seg_tract, by = c("locality" = "county", "tract" = "tract")) %>% 
  relocate(c("state", "locality", "tract"), .after = last_col())

# add tract names 
tract_data <- tract_data %>% 
  left_join(tractnames, by = c("GEOID" = "GEOID", "locality"= "locality", "tract"= "tract"), multiple = "all") %>%
  relocate(c("state", "locality", "tract"), .after = last_col())

# add tract names to block group
blkgrp_data <- blkgrp_data %>% 
  left_join(tractnames %>% select(-GEOID), by = c("locality"= "locality", "tract"= "tract"), multiple = "all") %>% 
  relocate(c("state", "locality", "tract", "blockgroup"), .after = last_col())

# Merge county data ----

# add life expectancy by county
county_data <- county_data %>% 
  left_join(life_exp, by = c("GEOID" = "GEOID", "locality"= "locality")) %>% 
  relocate(c("coname", "state", "locality"), .after = last_col())

# add segregation measures by county
county_data <- county_data %>% 
  left_join(seg_county, by = c("locality" = "county", "year" = "year")) %>% 
  relocate(c("coname", "state", "locality"), .after = last_col())

# add ALICE data by county
county_data <- county_data %>% 
  left_join(alice, by = c("GEOID" = "GEOID", "locality"= "locality", "coname" = "coname")) %>% 
  relocate(c("coname", "state", "locality"), .after = last_col())

```

## Add Geography 
```{r}
library(tigris)
options(tigris_use_cache = TRUE)
library(rmapshaper)

# get locality polygons
counties_geo <- counties(state = 'VA', year = tigeryr) %>% 
  subset(COUNTYFP %in% fips_codes) %>% 
  select(-NAME) %>% 
  ms_simplify(keep = .7, keep_shapes = TRUE) %>%
  sf::st_collection_extract()

# join coordinates to data
county_data_geo <- merge(county_data, counties_geo, by = "GEOID", duplicateGeoms = TRUE) # keep all obs (full_join)

# get tract polygons
tract_geo <- tracts(state = 'VA', county = fips_codes, year = tigeryr) %>% 
  rename(tr = NAME) %>% 
  ms_simplify(keep = .7, keep_shapes = TRUE) %>%
  sf::st_collection_extract()

# join coordinates to data
tract_data_geo <- merge(tract_data, tract_geo, by = c("GEOID"), duplicateGeoms = TRUE) # keep all obs (full_join)

# get block group polygons
blkgrp_geo <- block_groups(state = 'VA', county = fips_codes, year = tigeryr) %>%  # from tigris
  ms_simplify(keep = .7, keep_shapes = TRUE) %>%
  sf::st_collection_extract()

# join coordinates to data
blkgrp_data_geo <- merge(blkgrp_data, blkgrp_geo,  by = "GEOID", duplicateGeoms = TRUE) # keep all obs (full_join)

# combine into one data frame
all_data <- bind_rows("County" = county_data_geo, 
                      "Census Tract" = tract_data_geo,
                      "Block Group" = blkgrp_data_geo, 
                      .id = "GEO_LEVEL") %>% 
  select("GEOID","NAME","year","totalpopE","totalpopM",
         "whiteE","whiteM","blackE","blackM","asianE","asianM",
         "indigE","indigM","othraceE","othraceM","multiE","multiM","hispE","hispM",
         "povrateE","povrateM","cpovrateE","cpovrateM","hhincE","hhincM",  "hsmoreE","hsmoreM", "bamoreE",
         "bamoreM", "gradmoreE","gradmoreM","unempE","unempM",  "hlthinsE","hlthinsM","pubinsE","pubinsM",
         "earnE","earnM",   "disabledE","disabledM","giniE","giniM",   "vachseE","vachseM", "allhseE",
         "allhseM", "snapE","snapM",   "foreignbE","foreignbM","renter30E","renter30M","homeownE",
         "homeownM","vacrateE","vacrateM","perc_snaphseE","perc_snaphseM","perc_forbE","perc_forbM", "schlE",
         "schlM",   "age17E","age17M",  "age24E","age24M",  "age64E","age64M",  "age65E","age65M",
         "hhinc_blackE","hhinc_blackM","hhinc_asianE","hhinc_asianM","hhinc_multiE","hhinc_multiM",
         "hhinc_whiteE","hhinc_whiteM","hhinc_hispE","hhinc_hispM","lifeexpE","lifeexpM","lifeexp_hispE",
         "lifeexp_hispM","lifeexp_asianE","lifeexp_asianM","lifeexp_blackE","lifeexp_blackM","lifeexp_whiteE",
         "lifeexp_whiteM","dissim_wb","dissim_wh","inter_bw","inter_hw","iso_b","iso_h",   
         "poverty_households","alice_households","above_alice_households","alice_threshold_hh_under_65",   
         "alice_threshold_hh_65_years_and_over","per_poverty_households", 
         "per_alice_households","per_above_alice_households","coname", "state","locality", everything()   
  ) %>% 
  st_as_sf()

```

## Create attributes
```{r}

# match metadata to data
j <- match(metadata$varname, names(all_data))

# add label, sources and about to all_data
for(i in seq_along(j)){
  attr(all_data[[j[i]]], which = "label") <- metadata$label[i]
  attr(all_data[[j[i]]], which = "source") <- metadata$source[i]
  attr(all_data[[j[i]]], which = "about") <- metadata$about[i]
}

# create data frame of group and varnames
group_df <- metadata %>% 
  select(varname, group, label) %>% 
  filter(!is.na(group))
# categories <- unique(group_df$group)

# Create indicator lists based on geography
unique(all_data$GEO_LEVEL)
# [1] "County"       "Block Group"  "Census Tract"

# Block group
ind_bg <- all_data %>% 
  filter(GEO_LEVEL == "Block Group") %>%
  select(group_df$varname)  %>% 
  st_drop_geometry() %>% 
  map_lgl(~ !all(is.na(.x))) 

# Census tract
ind_ct <- all_data %>% 
  filter(GEO_LEVEL == "Census Tract") %>% 
  select(group_df$varname)  %>% 
  st_drop_geometry() %>% 
  map_lgl(~ !all(is.na(.x))) 

# Add indicator logicals to group_df and sort
# column bg - TRUE if variable available for Block Group
# column ct - TRUE if variable available for Census Tract
# all vars available for County

group_df <- cbind(group_df, ct = ind_ct, bg = ind_bg) %>% 
  arrange(group)

# Create cases for app interface
group_df <- group_df %>% 
  mutate(ind_name = case_when(ct == "FALSE" & bg == "FALSE" ~ paste0(label, " (County Only)"),
                              ct == "TRUE" & bg == "FALSE" ~ paste0(label, " (County & Tract)"),
                              ct == "TRUE" & bg == "TRUE" ~ paste0(label, " (All Levels)")),
         geo_level = case_when(ct == "FALSE" & bg == "FALSE" ~ "County",
                               ct == "TRUE" & bg == "FALSE" ~ "County, Census Tract",
                               ct == "TRUE" & bg == "TRUE" ~ "County, Census Tract, Block Group"))

k <- match(group_df$varname, names(all_data))

# add labels, sources and about to all_data
for(i in seq_along(k)){
  attr(all_data[[k[i]]], which = "geo_level") <- group_df$geo_level[i]
}

# one list of available indicators
ind_choices <- split(group_df, group_df$group) %>% 
  map(function(x)pull(x, varname, name = ind_name))

# create list of counties
counties <- levels(factor(county_data_geo$coname))
```

# App help features
```{r}
# get helpers
source('scripts/helpers.R')
```

# Define Color Palettes and Extras
```{r}
library(RColorBrewer)

nb.cols <- 10
mycolors <- colorRampPalette(brewer.pal(8, "YlGnBu"))(nb.cols)

# Create data dictionary for download
data_dict <- metadata %>% 
  unique() %>%
  select(-c("description")) %>%
  rename(variable_name = varname, description = label)

# Moved here from app.R ----
# was not deploying on shinyapps:
# https://stackoverflow.com/questions/61286108/error-in-cpl-transformx-crs-aoi-pipeline-reverse-ogrcreatecoordinatetrans
all_data$geometry <- st_transform(all_data$geometry, 4326)
all_data$pop <- as.character(all_data$totalpopE)
counties_geo$geometry <- st_transform(counties_geo$geometry, 4326)
# below variables for leaflet map boundary settings
bbox <- st_bbox(counties_geo) %>% as.vector()
cville_geo <- counties_geo %>% filter(NAMELSAD == "Charlottesville city")

fewpal <- c("#7DC462", "#0D95D0", "#E72F52", "#774FA0", "#EFB743", "#D44627")

# current variable that can't tercile at given geo levels
# For future developement: asianE and snapE can do median split at block level,
# but indigE, othraceE still have too many zeros at both tract and block for median split. 
# not available at census tract: indigE, othraceE; 
no_tercile_tract <- c("indigE", "othraceE")
# not avail at block group: indigE, othraceE, asianE, snapE 
no_tercile_block <- c("indigE", "othraceE", "asianE", "snapE")

```

# Save all data for app
```{r}
# Save for app ----
# create Rdata file
save(counties_geo, counties, all_data, mycolors,
     parks_sf, schools_sf, sabselem_sf, mcd_sf, group_df,
     ind_choices, helpers, data_dict, bbox, cville_geo, fewpal,
     no_tercile_tract, no_tercile_block,
     file = paste0(filepath_data, "app_data_5_2026.Rdata"))

# create data for app
library(qs2)
qs_save(list(counties_geo=counties_geo, counties=counties, all_data=all_data, mycolors=mycolors,
             parks_sf=parks_sf, schools_sf=schools_sf, sabselem_sf=sabselem_sf, mcd_sf=mcd_sf,
             group_df=group_df, ind_choices=ind_choices, helpers=helpers, data_dict=data_dict, bbox=bbox, 
             cville_geo=cville_geo, fewpal=fewpal, no_tercile_tract=no_tercile_tract, 
             no_tercile_block=no_tercile_block),
        file = paste0(filepath_app, "app_data_5_26.qs2"))

# app_data <- qs_read(paste0(filepath_app, "app_data_5_26.qs2"))
# 
# counties_geo <- app_data$counties_geo
# counties <-  app_data$counties
# all_data <- app_data$all_data
# mycolors <- app_data$mycolors
# parks_sf <- app_data$parks_sf
# schools_sf <- app_data$schools_sf
# sabselem_sf <- app_data$sabselem_sf
# mcd_sf <- app_data$mcd_sf
# group_df <- app_data$group_df
# ind_choices <- app_data$ind_choices
# helpers <- app_data$helpers
# data_dict <- app_data$data_dict
# bbox <- app_data$bbox
# cville_geo <- app_data$cville_geo
# fewpal <- app_data$fewpal
# no_tercile_tract <- app_data$no_tercile_tract
# no_tercile_block <- app_data$no_tercile_block

```



